One-Particle Density Matrix Polarization Susceptibility Tensors

### 2018-07-22 papers/1_dmatpol/paper-r2.tex
Second iteration of paper draft initiated.
New figure added to the paper.

In this work, the electric field\hyp{}induced change in the one\hyp{}electron density
has been expressed as a series of the one\hyp{}particle density matrix susceptibility tensors
interacting with electric fields. The analytical approximate expressions were derived
at the Hartree\hyp{}Fock theory, which served as a basis for construction of generalized
model that can be used for an arbitrary form of wavefunction as long as the one\hyp{}particle
density matrices are available. It is shown that it is possible to accurately predict the changes
in the one\hyp{}electron density of water molecule in spatially non\hyp{}uniform electric field distribution
generetated by point charges. The electric field\hyp{}induced interaction energies,
dipole moments and quadrupole moments
are quantitatively described by the present theory in weak and moderate electric fields (0.003 -- 0.02 a.u.).
%It is believed that the possibility of extension of the proposed model with high\hyp{}order susceptibilities
%opens new routes in quantum chemistry of fast and efficient calculations of strongly perturbed systems.
It is believed that the proposed model, after extensive development on its optimization, could open
new routes in quantum chemistry of fast and efficient calculations of strongly perturbed systems.
\end{abstract}

@@

The discovery of hybrid methods such as quantum mechanical\hyp{}molecular mechanical (QM/MM)
technique was a revolution in computational chemistry and allowed insightful theoretical studies
of quantum mechanical effects in biological systems because the high\hyp{}level
and costly quantum mechanical calculations
could be combined with low\hyp{}level and cheap classical dynamics.\cite{Warshel.Levitt.JMolBiol.1976,Senn.Thiel.Angew.2009}
%That was possible due to the fruitful incorporation of accurate \emph{ab initio} methods and a 
%semi\hyp{}classical notion of molecular motion combined with extensive experimental and semi\hyp{}empirical
%refinings.
Understanding physics that rules intermolecular interactions\cite{Jeziorski.Moszynski.Szalewicz.ChemRev.1994} 
helped to develop
a wide family of force fields, and served a basis for embedding Hamiltonian
within an effective environment and fragmentation methods.\cite{Gordon.Fedorov.Pruitt.Slipchenko.ChemRev.2012}
%The size of quantum level can be extended without compromising accuracy and maintaining good performance
%by using vast family of fragmentation methods, such as effective fragment potentials (EFP).
%The most difficult to tackle part of Hamiltonian can also be embedded within a widely understood effective environment
%leading to various embedding methodologies, realising similar goals.

However, even with the modern advances in computational chemistry, 
%accurate description of environmental effects
accurate description of phenomena due to complex heterogeneous molecular environments 
remains a challenge in the following areas: (i) describing the electronically excited states 
(electronic solvatochromism),\cite{Barbati.JACS.2014,
Szabla.Sponer.Jiri.Gora.JPCL.2015,
Bednarska.Zalesny.Tian.Murugan.Agren.Bartkowiak.Molecules.2017,
Jedrzejewska.Grabarz.Bartkowiak.Osmialowski.SpectChimActA.2018} 
(ii) understanding the vibrational properties of molecules in ground and electronically excited states
(vibrational or vibronic solvatochromisms).\cite{Blasiak.Londergan.Webb.Cho.ACR.2017} 
Interface between QM and MM levels is usually realized by electrostatic embedding
and only rarely includes the Pauli exclusion principle and dispersion.\cite{List.Olsen.Kongsted.PCCP.2016}
Unfortunately, it is quite difficult to efficiently treat the environment\hyp{}induced potential operator
because of complicated nature of intermolecular interactions, that theoretically
cannot be solely envisioned as a sum of electrostatic, dispersion and Pauli effects.
Perhaps the most frequently used efficient and accurate \emph{ab initio} fragmentation methods,
the effective fragment potential (EFP) methods, are limited to the ground\hyp{}state
chemistry at the Hartree\hyp{}Fock (HF)\cite{Gordon.Smith.Xu.Slipchenko.AnnuRevPhysChem.2013} 
or density functional theory (DFT) levels\cite{Nguyen.Pachter.Day.JCP.2014},
and, while they can approximately describe the environment composed of molecules in their ground states,
they cannot be used for highly correlated wavefunctions such as electronically excited states.
%Therefore, Hamiltonian embedding is often realised by Coulombic embedding with polarization
%of QM level, and only rarely the Pauli exclusion principle is taken into account. Although
%such approach is very accurate in certain cases, it is not guaranteed that it is sufficient
%for all cases.

Most of efforts to theoretically treat extended molecular aggregates are rooted in 
energy\hyp{} (or force field\hyp{})based philosophy. However,
in the past decade there has been considerable interest in density\hyp{}based approaches
of a wide scope and aims.\cite{Piquemal.Cisneros.Reinhardt.Gresh.Darden.JCP.2006,
Mandado.Hermida-Ramon.JCTC.2011,
Sun.Chan.ACR.2016,
Hedegard.Reiher.JCTC.2016} 
In particular, the importance of
one\hyp{}electron densities in chemistry is well known and acknowledged in the community,
though still not solved exactly.\cite{Holas.March.PhysRevA.1991}
%that seem to have an important advantage over the standard energetic approach.
With connection to the above, it would be incredibly insightful 
to establish a first\hyp{}principles connection bewteen the one\hyp{}particle
density and the perturbation in the environment through introducing the
\emph{density matrix susceptibility} (DMS) tensors. % because knowing the density distribution
%is equivalent to knowing the state of the molecule.
The main postulate of the present Work is that the associated change 
in the one\hyp{}particle density matrix, $\delta {\bf D}$, can be
deconvoluted into a product form of the generalized perturbation susceptibility ${\bf B}_i$
and the environment\hyp{}induced perturbation on a molecule, $\delta {\bf P}_i$,
evaluated at certain site $i$.
%and induced dipole moments $\delta {\BM\upmu}_i$ due to electric fields ${\bf E}_i$ 
%at the centroids of the occupied molecular orbitals,
%
\begin{equation*} \label{e:initial-model}
 \delta {\bf D} \approx \sum_i^{\text{Sites}} \text{[DMS of the $i$th site]} 
  \cdot  \delta {\bf P}_i
  %\rightarrow \sum_i^{\rm occ} {\bf B}^{(i)}_{\mu\nu} \cdot {\bf E}_i \;.
\end{equation*}
%
While $\delta {\bf P}_i$ contain critical information 
about the actual distortion of the electronic cloud,
perturbation susceptibility is assumed to be a property of unperturbed 
molecule and is to be precomputed only once. 
Thus, such DMS could be adapted into an efficient effective 
one\hyp{}electron fragment potential, in principle applicable to an arbitrary
level of theory, as long as the one\hyp{}particle densities are available.
It also means that such a technique would dramatically reduce costs of quantum chemistry calculations
becuase the task of performing full computation on a molecular aggregate
would reduce to much simpler matrix operations.
%
%a simple first\hyp{}principles model for prediction 
%of the perturbation\hyp{}induced one\hyp{}electron density
%change of structurally unrelaxed molecule is proposed here.

The simplest, yet perhaps the most relevant for chemistry, type of perturbation 
is the electrostatic perturbation.
It is worth mentioning that, at that moment, there is no simple \emph{ab initio}
method of computation of electron deformation densities without considering the entire molecular 
aggregate by using full quantum chemistry calculations.\cite{Horn.Yuezhi.Head-Gordon.JCP.2016,Horn.Head-Gordon.JCP.2015}
In this Work, we explore the susceptibility of the molecule's electronic density distribution
to polarize due to external electrostatic perturbation from first principles. 
The deformation density is expressed in terms
of the one\hyp{}particle density matrix polarization susceptibility tensors 
that are properties of isolated molecules and can be
extracted from quantum chemistry calculations. The focus on the density (matrix) instead of the energy
(scalar) is beneficial since it provides the directional tensor description
of the polarization that could be combined with other effects such as Pauli repulsion 
or dispersion,\cite{Mandado.Hermida-Ramon.JCTC.2011}
in contrast to the standard interaction energy approaches. We show that accurate polarization 
of the one\hyp{}particle density
matrix is achievable and can be realized as long as the one\hyp{}particle density matrices
are available. It is believed that the proposed method
could be a basis for a construction of an 
alternative approach in fragment\hyp{}based
modelling of extended molecular aggregates that is not limitted to the ground state chemistry.

@@

The electronic energy of a closed\hyp{}shell molecule is given by the time\hyp{}independent Schr{\"o}dinger equation,
%
\begin{equation}
 E^{(0)} = \tbraket{\Psi^{(0)}}{\mathscr{H}^{(0)}}{\Psi^{(0)}} \;,
\end{equation}
%
where $E$ stands for the total electronic energy, $\Psi$ the electronic wavefunction and $\mathscr{H}$
the Hamiltonian operator. In the above, the superscript $(0)$
denotes the reference or unperturbed state that is assumed to be known.
Under the influence of an external electrostatic perturbation, the total electronic energy
becomes
%
\begin{equation}
 E[{\bf F}({\bf r})] = \tbraket{\Psi}{\mathscr{H}^{(0)} + \mathscr{V}({\bf F}({\bf r}))}{\Psi}
\end{equation}
%
in which ${\bf F}({\bf r})$ is a non\hyp{}uniform electric field distribution.
The electric field\hyp{}induced energy change is an unknown functional 
of the associated change in the one\hyp{}electron density,
%
\begin{equation}\label{e:rho}
 \rho({\bf r}) = 2\sum_{\alpha\beta} D_{\alpha\beta} \phi_\alpha({\bf r}) \phi^*_\beta({\bf r}) \;.
\end{equation}
%
In Eq.~\eqref{e:rho} $\phi_\alpha({\bf r})$ is a basis funtion, which can be for instance 
an atomic orbital (AO) or a molecular orbital (MO), and $D_{\alpha\beta}$ is the corresponding
one\hyp{}particle density matrix.
If perturbation\hyp{}independent basis is chosen, the electric field\hyp{}induced change in the one\hyp{}particle density
is given by
%
\begin{equation}\label{e:rho-1}
 \delta \rho({\bf r}) = 2\sum_{\alpha\beta} \delta D_{\alpha\beta} \phi_\alpha({\bf r}) \phi^*_\beta({\bf r}) \;.
\end{equation}
%
Ideally, the electric field\hyp{}induced change in the ground state's one\hyp{}particle density 
follows its energy minimum of the perturbed state. In addition, all the perturbed 
excited state one\hyp{}electron densities must be excited state solutions of the Schr{\"o}dinger equation
with the exact energy\hyp{}minimized ground state.

Development of a first\hyp{}principles theory
that is capable of accurately capturing the electric field\hyp{}dependence of the one\hyp{}particle density
is extremely difficult. This is not only because of the very complicated (and generally unknown) solution
to the electronic correlation problem, but also because
a working model has to accurately predict all the $n^2$ elements of the density matrix
where $n$ is the size of basis set. Even relatively small errors
in a particular matrix element could accumulate tremendously when computing traces with other matrices,
and in turn, obstruct the evaluated properties with non\hyp{}negligible errors.
Therefore, we adopt the following approach: we first develop a qualitative first\hyp{}principles model 
at the Hartree\hyp{}Fock (HF)
level of theory, and then generalize it to the arbitrary level using the functional form that was found 
by considering HF level,
but treating the density matrix susceptibility tensors as cetrain functions of adjustable parameters further on.

@@

In the Hartree\hyp{}Fock theory,\cite{Roothaan.RevModPhys.1951} ${\bf D}$ is approximated as
%
\begin{equation}
 {\bf D} = {\bf C}{\bf C}^\dagger \;,
\end{equation}
%
where ${\bf C}$ describes the occupied molecular orbitals as linear combinations
of atomic orbitals (AO's), referred here to as the LCAO-MO matrix.
${\bf D}$ is an idempotent matrix and if expressed in orthogonal basis the following holds
%
\begin{equation}
 {\bf D}^2 = {\bf D} \;.
\end{equation}
%
The change in the molecular orbitals can be parameterized as suggested by McWeeny\cite{McWeeny.RevModPhys.1960}
by defining the idempotency\hyp{}preserving variation
%
\begin{equation}
 \delta {\bf C} = {\BM\Delta} \cdot {\bf C}^{(0)} \;,
\end{equation}
%
where $\BM\Delta$ is a square non\hyp{}singular and non\hyp{}symmetric matrix.
McWeeny showed that
%
\begin{equation} \label{e:dmat-change.exact}
 \delta {\bf D} = -{\bf D}^{(0)} + \left[ {\bf D}^{(0)} + {\bf v} \right]
                                   \left[ {\bf 1} + {\bf v}^\dagger{\bf v} \right]^{-1}
                                   \left[ {\bf D}^{(0)} + {\bf v}^\dagger \right] \;,
\end{equation}
%
where the auxiliary vector ${\bf v}$ is defined as
%
\begin{equation}
 {\bf v} \equiv \left[ {\bf 1} - {\bf D}^{(0)} \right] {\BM\Delta} {\bf D}^{(0)}  \;.
\end{equation}
%
Expanding Eq.~\eqref{e:dmat-change.exact} in a Taylor series around ${\bf v}={\bf 0}$ and
truncating it at linear terms with respect to ${\BM\Delta}$ gives
%
\begin{subequations} 
 \begin{align}
 \delta {\bf D} &\cong \left[ {\bf 1} - {\bf D}^{(0)} \right] {\BM\Delta} {\bf D}^{(0)} + 
                        {\bf D}^{(0)} {\BM\Delta}^\dagger \left[ {\bf 1} - {\bf D}^{(0)} \right]  
 \label{e:dD-1} \\  &= 
  \delta {\bf C}  {\bf C}^{(0)\dagger} + {\bf C}^{(0)}\delta {\bf C}^\dagger
           - {\bf D}^{(0)} \delta {\bf C}  {\bf C}^{{(0)}\dagger} - {\bf C}^{(0)}\delta {\bf C}^\dagger {\bf D}^{(0)} 
 \label{e:dD-2}
 \end{align}
\end{subequations}
%
The latter equality is true only if the basis set functions are orthogonal, i.e., 
${\bf C}^{(0)\dagger}{\bf C}^{(0)}={\bf 1}$,
which will be the case throughout the entire work, unless stated differently.
Substituting the result from Eq.~\eqref{e:dD-1}
into Eq.~\eqref{e:dmu-exact} leads to
%%
%\begin{equation} \label{e:delta-Taylor}
% \delta {\bf D} = \left[ {\bf v} + {\bf v}^\dagger \right]
%                + \left[ {\bf v}{\bf v}^\dagger - {\bf v}^\dagger{\bf v}\right] + \ldots 
%\end{equation}
%
\begin{multline} \label{e:dmu-4-exact.linear-approximation}
 \frac{1}{2} 
 \delta {\BM{\upmu}}
  \cong
   \Tr{ 
    \left[ 
         {\mathbb{M}} {\BM\Delta} {\bf D}^{(0)}  
    \right] }
%
  +\Tr{ 
    \left[ 
         {\mathbb{M}} {\bf D}^{(0)} {\BM\Delta}^{\dagger}
    \right] } \\
%
  -\Tr{ 
    \left[ 
         {\mathbb{M}} {\bf D}^{(0)} {\BM\Delta} {\bf D}^{(0)}
    \right] }
%
  -\Tr{ 
    \left[ 
         {\mathbb{M}} {\bf D}^{(0)} {\BM\Delta}^{\dagger} {\bf D}^{(0)}
    \right] } \;,
\end{multline}
%
It can be proven (see Appendix~\ref{a:orig-dep}) that such obtained
induced dipole moment is independent on the origin. 
This guarantees that the theory fulfills the total charge conservation requirement.
Therefore,
in the above result and also in later course of the Paper the dipole origin will be omitted. 
%where the dipole origin was omitted for notational clarity.

@@

The theory that is necessary to validate the density matrix susceptibility models from Section~\ref{s:2}
was implemented in our in\hyp{}house plugin to {\sc Psi4} quantum chemistry program.\cite{Psi4.JCTC.2017}
Water molecule, energy\hyp{}optimized by using the HF/6-311++G** 
method\cite{Roothaan.RevModPhys.1951,Krishnan.Binkley.Seeger.Pople.JCP.1980}, 
was selected as a test system which was subject to 
electrostatic perturbations due to random collections of point charges
generating non\hyp{}uniform electric fields. 
Cartesian coordinates of perturbing charges were drawn from uniform distribution of a set of fourty
three\hyp{}dimensional Euclidean points within a sphere of radius 6.0 Bohrs
and origin in the center of mass of water molecule.
In selection of random points, space inside the union of spheres surrounding
oxygen and hydrogen atoms with radii of 4.5 and 4.0 Bohrs, respectively, was
kept free from any point charges due to potential problems with the convergence of the 
self\hyp{}consistent field (SCF) calculations.
Charge values were selected from the one\hyp{}dimensional uniform distribution
of double precision numbers in the interval of $\vert q_{\rm max} \vert$, where $q_{\rm max}$
was tuned to obtain a desired ensemble of weak, moderate and strong electric field magnitudes
(with $q_{\rm max}=0.05$, $0.2$ and $0.4$ a.u., accordingly; cf. Table~\ref{t:fields}).
In this way, 100 random samples of 40 point charges were selected and exact solutions of HF\hyp{}SCF equations were 
performed for each validation procedure.

>>>>

In this work, the electric field\hyp{}induced change in the one\hyp{}electron density
has been expressed as a series of the one\hyp{}particle density matrix susceptibilities
interacting with electric fields. The analytical approximate expressions were derived
at the Hartree\hyp{}Fock theory, which served as a basis for construction of generalized
model that can be used for an arbitrary form of wavefunction as long as the one\hyp{}particle
density matrices are available. It is shown that it is possible to accurately predict the changes
in the one\hyp{}electron density of water molecule in spatially non\hyp{}uniform electric field distribution
generetated by point charges. The electric field\hyp{}induced interaction energies,
dipole moments and quadrupole moments
are quantitatively described by the present theory in weak and moderate electric fields (0.003 -- 0.02 a.u.).
%It is believed that the possibility of extension of the proposed model with high\hyp{}order susceptibilities
%opens new routes in quantum chemistry of fast and efficient calculations of strongly perturbed systems.
It is believed that the proposed model, after extensive development on its optimization, could open
new routes in quantum chemistry of fast and efficient calculations of strongly perturbed systems.
\end{abstract}

@@

The development of hybrid methods such as quantum mechanical\hyp{}molecular mechanical (QM/MM)
technique was a revolution in computational chemistry and allowed insightful theoretical studies
of quantum mechanical effects in biological systems because the high\hyp{}level
and costly quantum mechanical calculations
could be combined with low\hyp{}level and cheap classical dynamics.\cite{Warshel.Levitt.JMolBiol.1976,Senn.Thiel.Angew.2009}
%That was possible due to the fruitful incorporation of accurate \emph{ab initio} methods and a 
%semi\hyp{}classical notion of molecular motion combined with extensive experimental and semi\hyp{}empirical
%refinings.
Understanding physics that rules intermolecular interactions\cite{Jeziorski.Moszynski.Szalewicz.ChemRev.1994} 
helped to develop
a wide family of force fields, and served as a basis for embedding Hamiltonian
within an effective environment and fragmentation methods.\cite{Gordon.Fedorov.Pruitt.Slipchenko.ChemRev.2012}
%The size of quantum level can be extended without compromising accuracy and maintaining good performance
%by using vast family of fragmentation methods, such as effective fragment potentials (EFP).
%The most difficult to tackle part of Hamiltonian can also be embedded within a widely understood effective environment
%leading to various embedding methodologies, realising similar goals.

However, even with the modern advances in computational chemistry, 
%accurate description of environmental effects
accurate description of phenomena due to complex heterogeneous molecular environments 
remains a challenge in the following areas: (i) describing the electronically excited states 
(electronic solvatochromism),\cite{Barbati.JACS.2014,
Szabla.Sponer.Jiri.Gora.JPCL.2015,
Bednarska.Zalesny.Tian.Murugan.Agren.Bartkowiak.Molecules.2017,
Jedrzejewska.Grabarz.Bartkowiak.Osmialowski.SpectChimActA.2018} 
(ii) understanding the vibrational properties of molecules in ground and electronically excited states
(vibrational or vibronic solvatochromisms).\cite{Blasiak.Londergan.Webb.Cho.ACR.2017} 
Interface between QM and MM levels is usually realized by electrostatic embedding
and only rarely includes the Pauli exclusion principle and dispersion.\cite{List.Olsen.Kongsted.PCCP.2016}
Unfortunately, it is quite difficult to efficiently treat the environment\hyp{}induced potential operator
because of complicated nature of intermolecular interactions, that theoretically
cannot be solely envisioned as a sum of electrostatic, dispersion and Pauli effects.
Perhaps the most frequently used efficient and reasonably accurate \emph{ab initio} fragmentation methods,
the effective fragment potential (EFP) methods, are limited to the ground\hyp{}state
chemistry at the Hartree\hyp{}Fock (HF)\cite{Gordon.Smith.Xu.Slipchenko.AnnuRevPhysChem.2013} 
or density functional theory (DFT) levels\cite{Nguyen.Pachter.Day.JCP.2014}.
Unfortunately, while they can approximately describe the environment composed of molecules in their ground states,
they cannot be used for highly correlated wavefunctions such as electronically excited states.
%Therefore, Hamiltonian embedding is often realised by Coulombic embedding with polarization
%of QM level, and only rarely the Pauli exclusion principle is taken into account. Although
%such approach is very accurate in certain cases, it is not guaranteed that it is sufficient
%for all cases.

Most of efforts to theoretically treat extended molecular aggregates are rooted in 
energy\hyp{} (or force field\hyp{})based philosophy. However,
in the past decade there has been considerable interest in density\hyp{}based approaches
of a wide scope and aims.\cite{Piquemal.Cisneros.Reinhardt.Gresh.Darden.JCP.2006,
Mandado.Hermida-Ramon.JCTC.2011,
Sun.Chan.ACR.2016,
Hedegard.Reiher.JCTC.2016} 
In particular, the importance of
one\hyp{}electron densities in chemistry is well known and acknowledged in the community,
though still not solved exactly.\cite{Holas.March.PhysRevA.1991}
%that seem to have an important advantage over the standard energetic approach.
With connection to the above, it would be incredibly insightful 
to establish a first\hyp{}principles connection bewteen the one\hyp{}particle
density and the perturbation in the environment through introducing the
\emph{density matrix susceptibility} (DMS) tensors. % because knowing the density distribution
%is equivalent to knowing the state of the molecule.
The main postulate of the present work is that the associated change 
in the one\hyp{}particle density matrix, $\delta {\bf D}$, can be
represented in a product form of the generalized perturbation susceptibility ${\bf B}_i$
and the environment\hyp{}induced perturbation on a molecule, $\delta {\bf P}_i$,
evaluated at certain site $i$.
%and induced dipole moments $\delta {\BM\upmu}_i$ due to electric fields ${\bf E}_i$ 
%at the centroids of the occupied molecular orbitals,
%
\begin{equation*} \label{e:initial-model}
 \delta {\bf D} \approx \sum_i^{\text{Sites}} \text{[DMS of the $i$th site]} 
  \cdot  \delta {\bf P}_i
  %\rightarrow \sum_i^{\rm occ} {\bf B}^{(i)}_{\mu\nu} \cdot {\bf E}_i \;.
\end{equation*}
%
While $\delta {\bf P}_i$ contains critical information 
about the actual distortion of the electronic cloud,
perturbation susceptibility is assumed to be a property of unperturbed 
molecule and is to be precomputed only once. 
Thus, such DMS could be adapted into an efficient effective 
one\hyp{}electron fragment potential, in principle applicable to an arbitrary
level of theory, as long as the one\hyp{}particle densities are available.
It also means that such a technique would dramatically reduce costs of quantum chemistry calculations
becuase the task of performing full computation on a molecular aggregate
would reduce to much simpler matrix operations.
%
%a simple first\hyp{}principles model for prediction 
%of the perturbation\hyp{}induced one\hyp{}electron density
%change of structurally unrelaxed molecule is proposed here.

The simplest, yet perhaps the most relevant for chemistry, type of perturbation 
is the electrostatic perturbation.
It is worth mentioning that, at that moment, there is no simple \emph{ab initio}
method of computation of electron deformation densities without considering the entire molecular 
aggregate by using full quantum chemistry calculations.\cite{Horn.Yuezhi.Head-Gordon.JCP.2016,Horn.Head-Gordon.JCP.2015}
In this work, we explore the susceptibility of the molecule's electronic density distribution
to polarize due to external electrostatic perturbation from first principles. 
The deformation density is expressed in terms
of the one\hyp{}particle density matrix polarization susceptibility tensors 
that are properties of isolated molecules and can be
extracted from quantum chemistry calculations. The focus on the density (matrix) instead of the energy
(scalar) is beneficial since it provides the directional tensor description
of the polarization that could be combined with other effects such as Pauli repulsion 
or dispersion,\cite{Mandado.Hermida-Ramon.JCTC.2011}
in contrast to the standard interaction energy approaches. We show that accurate polarization 
of the one\hyp{}particle density
matrix is achievable and can be realized as long as the one\hyp{}particle density matrices
are available. It is believed that the proposed method
could be a basis for a construction of an 
alternative approach in fragment\hyp{}based
modelling of extended molecular aggregates that is not limitted to the ground state chemistry.

@@

The electronic energy of a closed\hyp{}shell molecule is given by the time\hyp{}independent Schr{\"o}dinger equation,
%
\begin{equation}
 E^{(0)} = \tbraket{\Psi^{(0)}}{\mathscr{H}^{(0)}}{\Psi^{(0)}} \;,
\end{equation}
%
where $E$ stands for the total electronic energy, $\Psi$ the electronic wavefunction and $\mathscr{H}$
the Hamiltonian operator. In the above, the superscript $(0)$
denotes the reference or unperturbed state that is assumed to be known.
Under the influence of an external electrostatic perturbation, the total electronic energy
becomes
%
\begin{equation}
 E[{\bf F}({\bf r})] = \tbraket{\Psi}{\mathscr{H}^{(0)} + \mathscr{V}({\bf F}({\bf r}))}{\Psi}
\end{equation}
%
in which ${\bf F}({\bf r})$ is a non\hyp{}uniform electric field distribution.
The electric field\hyp{}induced energy change is an unknown functional 
of the associated change in the one\hyp{}electron density,
%
\begin{equation}\label{e:rho}
 \rho({\bf r}) = 2\sum_{\alpha\beta} D_{\alpha\beta} \phi_\alpha({\bf r}) \phi^*_\beta({\bf r}) \;.
\end{equation}
%
In Eq.~\eqref{e:rho} $\phi_\alpha({\bf r})$ is a basis funtion, which can be for instance 
an atomic orbital (AO) or a molecular orbital (MO), and $D_{\alpha\beta}$ is the corresponding
one\hyp{}particle density matrix.
If perturbation\hyp{}independent basis is chosen, the electric field\hyp{}induced change in the one\hyp{}particle density
is given by
%
\begin{equation}\label{e:rho-1}
 \delta \rho({\bf r}) = 2\sum_{\alpha\beta} \delta D_{\alpha\beta} \phi_\alpha({\bf r}) \phi^*_\beta({\bf r}) \;.
\end{equation}
%
The electric field\hyp{}induced change in the ground state's one\hyp{}particle density 
follows its energy minimum of the perturbed state. In addition, all the perturbed 
excited state one\hyp{}electron densities must be excited state solutions of the Schr{\"o}dinger equation
with the exact energy\hyp{}minimized ground state. 
In the limit of a weak and spatially uniform electric field,
the polarization\hyp{}induced density matrix change is given by the follwoing Taylor expansion
%
\begin{equation}\label{e:dD-Taylor}
 \delta D_{\alpha\beta} = \frac{\partial D_{\alpha\beta}}{\partial {\bf F}}\Bigg|_{{\bf F}={\bf 0}}  \cdot {\bf F} 
   + \frac{1}{2} 
     \frac{\partial^2 D_{\alpha\beta}}{\partial {\bf F} \otimes \partial {\bf F}}\Bigg|_{{\bf F}={\bf 0}} : {\bf F} \otimes {\bf F}
   + \ldots 
\end{equation}
%
In the above expression, the density matrix susceptibilities are simply equal to the derivatives of the density matrix
with respect to electric field (up to a multiplicative constant).

In a general case, i.e., when electric field is non\hyp{}uniform, the development of a first\hyp{}principles theory
that is capable of accurately capturing the electric field\hyp{}dependence of the one\hyp{}particle density
becomes extremely difficult. This is not only because of the very complicated (and generally unknown) solution
to the electronic correlation problem, but also because
a working model has to accurately predict all the $n^2$ elements of the density matrix
where $n$ is the size of basis set. Even relatively small errors
in a particular matrix element could accumulate tremendously when performing two\hyp{}
and four\hyp{}index contractions,
and in turn, obstruct the evaluated properties with non\hyp{}negligible errors.
Single\hyp{}centered expansions, like the one from Eq.~\eqref{e:dD-Taylor}, although valid for 
an arbitrary level of theory,
are probably not sufficient enough to accurately describe the local polarization
of electron density distribution when the electric fields are highly non\hyp{}uniform in space.
Due to all of these reasons, we adopt the following approach: 
we first develop a qualitative first\hyp{}principles distributed site model at the Hartree\hyp{}Fock (HF)
level of theory, and then generalize it to the arbitrary level using the functional form that was found 
by considering HF level,
but treating the density matrix susceptibility tensors as cetrain functions of adjustable parameters further on.

@@

In the Hartree\hyp{}Fock theory,\cite{Roothaan.RevModPhys.1951} ${\bf D}$ is approximated as
%
\begin{equation}
 {\bf D} = {\bf C}{\bf C}^\dagger \;,
\end{equation}
%
where ${\bf C}$ describes the occupied molecular orbitals as linear combinations
of atomic orbitals (AO's), referred here to as the LCAO-MO matrix.
${\bf D}$ is an idempotent matrix and if expressed in orthogonal basis the following holds
%
\begin{equation}
 {\bf D}^2 = {\bf D} \;.
\end{equation}
%
The change in the molecular orbitals can be parameterized as suggested by McWeeny\cite{McWeeny.RevModPhys.1960}
by defining the idempotency\hyp{}preserving variation
%
\begin{equation}
 \delta {\bf C} = {\BM\Delta} \cdot {\bf C}^{(0)} \;,
\end{equation}
%
where $\BM\Delta$ is a square non\hyp{}singular and non\hyp{}symmetric matrix.
McWeeny showed that
%
\begin{equation} \label{e:dmat-change.exact}
 \delta {\bf D} = -{\bf D}^{(0)} + \left[ {\bf D}^{(0)} + {\bf v} \right]
                                   \left[ {\bf 1} + {\bf v}^\dagger{\bf v} \right]^{-1}
                                   \left[ {\bf D}^{(0)} + {\bf v}^\dagger \right] \;,
\end{equation}
%
where the auxiliary vector ${\bf v}$ is defined as
%
\begin{equation}
 {\bf v} \equiv \left[ {\bf 1} - {\bf D}^{(0)} \right] {\BM\Delta} {\bf D}^{(0)}  \;.
\end{equation}
%
Expanding Eq.~\eqref{e:dmat-change.exact} in a Taylor series around ${\bf v}={\bf 0}$ and
truncating it at linear terms with respect to ${\BM\Delta}$ gives
%
\begin{subequations} 
 \begin{align}
 \delta {\bf D} &\cong \left[ {\bf 1} - {\bf D}^{(0)} \right] {\BM\Delta} {\bf D}^{(0)} + 
                        {\bf D}^{(0)} {\BM\Delta}^\dagger \left[ {\bf 1} - {\bf D}^{(0)} \right]  
 \label{e:dD-1} \\  &= 
  \delta {\bf C}  {\bf C}^{(0)\dagger} + {\bf C}^{(0)}\delta {\bf C}^\dagger
           - {\bf D}^{(0)} \delta {\bf C}  {\bf C}^{{(0)}\dagger} - {\bf C}^{(0)}\delta {\bf C}^\dagger {\bf D}^{(0)} 
 \label{e:dD-2}
 \end{align}
\end{subequations}
%
The latter equality is true only if the basis set functions are orthogonal, i.e., 
${\bf C}^{(0)\dagger}{\bf C}^{(0)}={\bf 1}$,
which will be the case throughout the entire work, unless stated otherwise.
Substituting the result from Eq.~\eqref{e:dD-1}
into Eq.~\eqref{e:dmu-exact} leads to
%%
%\begin{equation} \label{e:delta-Taylor}
% \delta {\bf D} = \left[ {\bf v} + {\bf v}^\dagger \right]
%                + \left[ {\bf v}{\bf v}^\dagger - {\bf v}^\dagger{\bf v}\right] + \ldots 
%\end{equation}
%
\begin{multline} \label{e:dmu-4-exact.linear-approximation}
 \frac{1}{2} 
 \delta {\BM{\upmu}}
  \cong
   \Tr{ 
    \left[ 
         {\mathbb{M}} {\BM\Delta} {\bf D}^{(0)}  
    \right] }
%
  +\Tr{ 
    \left[ 
         {\mathbb{M}} {\bf D}^{(0)} {\BM\Delta}^{\dagger}
    \right] } \\
%
  -\Tr{ 
    \left[ 
         {\mathbb{M}} {\bf D}^{(0)} {\BM\Delta} {\bf D}^{(0)}
    \right] }
%
  -\Tr{ 
    \left[ 
         {\mathbb{M}} {\bf D}^{(0)} {\BM\Delta}^{\dagger} {\bf D}^{(0)}
    \right] } \;,
\end{multline}
%
It can be proven (see Appendix~\ref{a:orig-dep}) that such obtained
induced dipole moment is independent on the origin. 
This guarantees that the theory fulfills the total charge conservation requirement.
Therefore,
in the above result and also in later course of the paper the dipole origin will be omitted. 
%where the dipole origin was omitted for notational clarity.

@@

The theory that is necessary to validate the density matrix susceptibility models from Section~\ref{s:2}
was implemented in our in\hyp{}house plugin to {\sc Psi4} quantum chemistry program.\cite{Psi4.JCTC.2017}
Water molecule, energy\hyp{}optimized by using the HF/6-311++G** 
method\cite{Roothaan.RevModPhys.1951,Krishnan.Binkley.Seeger.Pople.JCP.1980}, 
was selected as a test system which was subject to 
electrostatic perturbations due to randomly selected uniform electric fields and random collections of point charges
generating non\hyp{}uniform electric fields. 
Components of spatially uniform electric fields were determined 
from the uniform distribution in the interval of $\vert 0.02 \vert$ a.u.
generating average field of 0.035 a.u. in an ensemble of 100 samples.
Cartesian coordinates of perturbing charges were drawn from the uniform distribution of a set of fourty
three\hyp{}dimensional Euclidean points within a sphere of radius 6.0 Bohrs
and origin in the center of mass of water molecule.
In selection of random points, space inside the union of spheres surrounding
oxygen and hydrogen atoms with radii of 4.5 and 4.0 Bohrs, respectively, was
kept free from any point charges due to potential problems with the convergence of the 
self\hyp{}consistent field (SCF) calculations.
Electric charge values were selected from the one\hyp{}dimensional uniform distribution
of double precision numbers in the interval of $\vert q_{\rm max} \vert$, where $q_{\rm max}$
was tuned to obtain a desired ensemble of weak, moderate and strong electric field magnitudes
(with $q_{\rm max}=0.05$, $0.2$ and $0.4$ a.u., accordingly; cf. Table~\ref{t:fields}).
In this way, independent sets of 100 random samples of uniform electric fields as well as 
40 point charges were selected and exact solutions of HF\hyp{}SCF equations were 
performed for each validation procedure.
